Computational methods in landscape ecology

### 2024-08-29 Sections/02-spatial-patterns.qmd
Adding comments to spatial patterns section

Moreover, many other entropy-based metrics have been proposed for use in landscape ecology.
The Renyi [@Renyi1961] and Gibbs entropies <!-- REF -->, which are both generalizations of the Shannon entropy, are applied to quantify the complexity of the landscape.
The Rao quadratic entropy [@Rao1982] has also been applied recently [@Rocchini2017], as it measures not only the relative abundances of elements, but also the pairwise dissimilarities or distances between them.
Thus, it can be useful in cases, when the dissimilarities between LULC types, and not only their categories, are relevant.
Another recent development is the use of Kullback-Leibler divergence (also known as relative entropy), which is a measure of differences between two probability distributions to describe patterns across scales [@Huckeba2023].

@@

<!-- https://doi.org/10.1371/journal.pone.0304215 -->
Despite the extensive use of the landscape mosaic model and the landscape metrics based on it, major limitations have been identified regarding both its ability to quantify the spatial structure of landscapes and their relationship to ecological processes [@Gustafson1998; @Li2004b; @Dramstad2009; @Kupfer2012; @Lausch2015; @Frazier2017].
<!-- two key considerations to select metrics: (i) ability to consistently identify structural landscape components and (ii) ability to explain relationships with ecological processes (ecological relevance), e.g., biodiversity, often referred to as functional metrics - the challenge of matching the scales of observation to the ecological scales affecting organisms or processes of interest, i.e., the characteristic scales (from "Spatial dependence of predictions from image segmentation: a variogram-based method to determine appropriate scales for producing land-management information") -->
<!-- Here we focus on the first issue, namely the ability of landscape metrics to consistently capture the spatial structure of landscapes. -->
<!-- Key issues of structural metrics [@Dramstad2009]: sensitivity to scale, i.e., resolution and extent, sensitivity to the choice of LULC scheme, redundancy and interpretation. Let us go through them in a different order: -->
In line with Dramstad [@Dramstad2009], we review four major issues of landscape metrics, namely their sensitivity to both spatial scale and thematic resolution, redundancy and interpretation.

@@

<!-- (i) scale: https://doi.org/10.1007/s10980-014-0066-3 "Behavior of class-level landscape metrics across gradients of class aggregation and area"-->
<!-- "identifying the characteristic scales of relevant patterns and processes is a critical first step in designing a successful research project" [@Wu2007, pp. 119]. -->
Identifying the characteristic scale at which the organisms or processes of interest operate is a critical first step towards establishing sound relationships between pattern and process [@Levin1992; @Wu2007; @Cushman2008Landscape].
To that end, a key requirement it to understand the sensitivity of landscape metrics to spatial scale in terms of both resolution (grain) and extent.

A key concept to consider regarding grain size is the minimum mapping unit (MMU), namely the smallest area entity that can be mapped and analyzed as a discrete patch.
<!-- Accordingly, a grain size several times smaller than the smallest patch in order to grasp (even schematically) its shape. -->
Since the input maps should represent not only the location of patches but also their shape, a grain size several times smaller than the smallest patch (i.e., the MMU) in order to grasp (even schematically) its shape.
Even though the MMU is often assumed as a natural characteristic of the dataset, it has an unsurprising impact on the computed landscape metrics, especially in terms of the number of patches and their size distribution [@Saura2002].
<!-- On the other hand, landscape metrics such as the landscape division index, largest patch index and area-weighted mean shape indices show more predictable responses to changing the MMU. -->

<!-- Since the applications of landscape metrics are manifold, the choices of grain and especially extent vary largely depending on the context of the study. -->
<!-- The growing availability of global remote sensing products at homogenized resolution has resulted in a myriad of studies sharing the same resolution, e.g., based on Landsat scenes at 30 m. At the same time, the choice of spatial extent varies largely depending on the phenomenon of interest, e.g., from large transects of forests in natural parks to small buffers along the urban-rural gradient. -->
While the choice of grain is often constrained by the available data, e.g., Landsat scenes at 30 m resolution, the selection of spatial extent and the way that it is sampled varies largely depending on the phenomenon of interest, e.g., from large transects of forests in natural parks to small buffers along the urban-rural gradient.
At the same time, the choice of spatial extent and sampling strategy is often subject to many arbitrary criteria that can largely influence the ultimate results. Such an issue has long been recognized in physical geography as the modifiable areal unit problem (MAUP) [@Openshaw1979; @Openshaw1984].
<!-- The MAUP has two interrelated components, namely the scale problem and the zoning problem -->
<!-- A first approach to dealing with the MAUP is to identify individual entities that are ecologically meaninful and not modifiable [@Jelinski1996]. In the landscape matrix model, such entities could be target patches of interest. However, it is not always possible to identify basic entities, especially since ecological processes of interest typically involve the integration of spatially distributed data. -->
Approaches to dealing with the MAUP in the context of landscape metrics include the identification of individual entities that are ecologically meaningful and not modifiable, the evaluation of "optimal" zoning systems that maximize interzonal variations and minimize intrazonal variation and the analysis of the sensitivity of the zoning configuration [@Jelinski1996; @Duque2018; @Fletcher2023].
<!-- TODO: use a moving window measurement device to "postpone" the MAUP [@Riitters2019] -->

TODO: section about approaches to sample landscapes to compute metrics and available software tools, e.g., zonal analysis with polygons (e.g., administrative limits), grids, transects or buffers along a landscape gradient, moving windows [@Bosch2019; @Hesselbarth2019; @Huais2024]?

<!-- While selecting the adequate scale should be based on the phenomenon of interest, researchers are often constrained by the availability of data. -->
<!-- Therefore, a thorough understanding of these relationships is essential in order to compare spatial patterns measured at different scales. -->
<!-- Several types of responses of landscape metrics to changing scale have been identified in computational experiments by changing both the spatial resolution (grain) and extent. -->
The scale effects of landscape metrics can be assessed by computing scalograms, i.e., the response curves of metrics to changing scales. For instance, some metrics related to the number and density of patches and edges show consistent linear and power-law scaling relationships to changing grain across a wide diversity of simulated and real landscapes, whereas other metrics exhibit less predictable responses [@Turner1989b; @Saura2001; @Wu2002a; @Wu2004; @Saura2007; @Simova2012; @Arganaraz2014; @Frazier2021].
<!-- Advancing the science of scaling in landscape ecology is key to "permit the prediction and inference of quantities measured at one scale to another" [@Frazier2023] -->
<!-- also multi-scale is key to habitat assessment, see "Multi-scale habitat selection modeling: a review and outlook" -->
<!-- Predictable scaling relationships can be exploited to estimate spatial patterns at higher resolutions -->
Nonetheless, the effects of changing the sampling extent can be less predictable than those of changing grain size [@Wu2004].
For instance, in landscapes with small extent to grain ratio, boundary effects are likely to dominate and truncate patches, resulting in unreliable landscape metric computations.
In order to address scale issues and advance the science of scaling in landscape ecology, studies should explicitly consider the concept of scope, namely the ratio of extent to resolution [@Schneider2001].
As suggested by Frazier [@Frazier2023], this can be done following three key steps, namely by considering scope guidelines when designing experiments, explicitly reporting grain and extent so that scope can be calculated allowing quantitative comparability with other studes and undertaking comparisons and replications based on scope.
The latter is showcased in a proof of concept example experiment, where landscapes from the same ecoregion are sampled using different grain and extent and then the distributions of the computed landscape metrics are visualized through ridgeline plots colored by scope [@Frazier2023].

TODO: section about landscape metric rescaling [@Saura2007; @Arganaraz2014; @Frazier2014; @Frazier2016; @Frazier2021]?

@@

TODO: mention hierarchichal patch dynamics [@Wu1999]?

<!-- (ii) choice of LULC scheme/thematic resolution and image processing/classification procedure: https://doi.org/10.1016/j.foreco.2009.12.016 https://doi.org/10.1016/j.isprsjprs.2007.03.004 https://doi.org/10.1007/s10021-005-0119-1 https://doi.org/10.1007/s10980-006-9035-9 https://doi.org/10.1007/s10980-022-01546-1 -->
<!-- Scale and heterogeneity are inherently linked (Wu 2007), so it is not surprising that the results indicate downscaling accuracy varies with heterogeneity. -->
TODO: sensitivity to thematic resolution [@Huang2006; @Bailey2007; @Buyantuyev2007; @Castilla2009].

@@

@@

Many landscape ecology studies are based on the raster data model.
However, the decision on the data model is often driven by data and software availability, or the familiarity with the approach [@Costanza2019].
<!-- Most LULC maps are provided as raster as they often related to underlying remote sensing products. -->
For example, several early landscape ecology studies were based on polygonal vector data that was converted to raster data. <!-- REF -->
Yet, nowadays, vector data is increasingly available, and most geospatial software supports it.

>>>>

Moreover, many other entropy-based metrics have been proposed for use in landscape ecology.
The Renyi [@Renyi1961] and Gibbs entropies <!--[REF]-->, which are both generalizations of the Shannon entropy, are applied to quantify the complexity of the landscape.
The Rao quadratic entropy [@Rao1982] has also been applied recently [@Rocchini2017], as it measures not only the relative abundances of elements, but also the pairwise dissimilarities or distances between them.
Thus, it can be useful in cases, when the dissimilarities between LULC types, and not only their categories, are relevant.
Another recent development is the use of Kullback-Leibler divergence (also known as relative entropy), which is a measure of differences between two probability distributions to describe patterns across scales [@Huckeba2023].

@@

<!-- MH: Is the landscape mosaic after e.g., Vogt et al. 10.1371/journal.pone.0304215 the same as the patch mosaic model and landscape metrics? I think there is some confusion here. We already discuss metrics above. I think it would be more valuable to discuss the LLM and the most recent additions after Vogt et al. here? -->

<!-- https://doi.org/10.1371/journal.pone.0304215 -->
Despite the extensive use of the landscape mosaic model and metrics based on it, major limitations have been identified regarding both its ability to quantify the spatial structure of landscapes and their relationship to ecological processes [@Gustafson1998; @Li2004b; @Dramstad2009; @Kupfer2012; @Lausch2015; @Frazier2017].
<!-- two key considerations to select metrics: (i) ability to consistently identify structural landscape components and (ii) ability to explain relationships with ecological processes (ecological relevance), e.g., biodiversity, often referred to as functional metrics - the challenge of matching the scales of observation to the ecological scales affecting organisms or processes of interest, i.e., the characteristic scales (from "Spatial dependence of predictions from image segmentation: a variogram-based method to determine appropriate scales for producing land-management information") -->
<!-- Here we focus on the first issue, namely the ability of landscape metrics to consistently capture the spatial structure of landscapes. -->
<!-- Key issues of structural metrics [@Dramstad2009]: sensitivity to scale, i.e., resolution and extent, sensitivity to the choice of LULC scheme, redundancy and interpretation. Let us go through them in a different order: -->
Here, we review four major issues of landscape metrics, namely their sensitivity to both spatial scale and thematic resolution, redundancy and interpretation [@Dramstad2009].

@@

<!-- MH: Suggestion is to focus much more on computational issues here. Currently, this is more about landscape metrics and their behavior in general. I like the point about growing availability of remote sensing products. Maybe we can focus a bit more on that and cut a lot of the more general, theoretical behavior parts? -->

<!-- (i) scale: https://doi.org/10.1007/s10980-014-0066-3 "Behavior of class-level landscape metrics across gradients of class aggregation and area"-->
<!-- "identifying the characteristic scales of relevant patterns and processes is a critical first step in designing a successful research project" [@Wu2007, pp. 119]. -->
Identifying scales at which the ecological processes of interest operate is a critical first step towards establishing relationships between pattern and process [@Levin1992; @Wu2007; @Cushman2008Landscape].
To that end, a key requirement is to understand the sensitivity of landscape metrics to spatial scale in terms of both grain and extent.

A key concept to consider regarding grain size is the minimum mapping unit (MMU), namely the smallest area entity that can be mapped and analyzed as a discrete patch.
<!-- Accordingly, a grain size several times smaller than the smallest patch in order to grasp (even schematically) its shape. -->
Since the input maps should represent not only the location of patches but also their shape, the grain size should be several times smaller than the smallest patch (i.e., the MMU) in order to quantify shapes.
Even though the MMU is often assumed as a natural characteristic of the dataset, it has an impact on the computed landscape metrics, especially on the number of patches and their size distribution [@Saura2002].
<!-- On the other hand, landscape metrics such as the landscape division index, largest patch index and area-weighted mean shape indices show more predictable responses to changing the MMU. -->

<!-- Since the applications of landscape metrics are manifold, the choices of grain and especially extent vary largely depending on the context of the study. -->
<!-- The growing availability of global remote sensing products at homogenized resolution has resulted in a myriad of studies sharing the same resolution, e.g., based on Landsat scenes at 30 m. At the same time, the choice of spatial extent varies largely depending on the phenomenon of interest, e.g., from large transects of forests in natural parks to small buffers along the urban-rural gradient. -->
While the choice of grain is often constrained by the available data, e.g., Landsat scenes at 30 m resolution, the selection of spatial extent varies largely depending on the phenomenon of interest, e.g., from large transects of forests in natural parks to small buffers along the urban-rural gradient. <!-- REF -->
The spatial extent and sampling strategy is often subject to many arbitrary criteria that can largely influence results. This issue has long been recognized in physical geography as the modifiable areal unit problem [MAUP; @Openshaw1979; @Openshaw1984].
<!-- The MAUP has two interrelated components, namely the scale problem and the zoning problem -->
<!-- A first approach to dealing with the MAUP is to identify individual entities that are ecologically meaninful and not modifiable [@Jelinski1996]. In the landscape matrix model, such entities could be target patches of interest. However, it is not always possible to identify basic entities, especially since ecological processes of interest typically involve the integration of spatially distributed data. -->
Dealing with the MAUP in the context of landscape metrics include the identification of individual entities that are ecologically meaningful and not modifiable, the evaluation of "optimal" zoning systems that maximize interzonal variations and minimize intrazonal variation, and the analysis of the sensitivity of the zoning configuration [@Jelinski1996; @Duque2018; @Fletcher2023].
<!-- TODO: use a moving window measurement device to "postpone" the MAUP [@Riitters2019] -->

<!-- MH: Can we move this to the sampling section maybe? -->
<!-- TODO: section about approaches to sample landscapes to compute metrics and available software tools, e.g., zonal analysis with polygons (e.g., administrative limits), grids, transects or buffers along a landscape gradient, moving windows [@Bosch2019; @Hesselbarth2019; @Huais2024]? -->

<!-- While selecting the adequate scale should be based on the phenomenon of interest, researchers are often constrained by the availability of data. -->
<!-- Therefore, a thorough understanding of these relationships is essential in order to compare spatial patterns measured at different scales. -->
<!-- Several types of responses of landscape metrics to changing scale have been identified in computational experiments by changing both the spatial resolution (grain) and extent. -->
The scale effects of landscape metrics can be assessed by computing scalograms, i.e., the response curves of metrics to changing scales. 
Some metrics related to the number of patches, edges show consistent linear and power-law scaling relationships to changing grain, while other metrics exhibit less predictable responses [@Turner1989b; @Saura2001; @Wu2002a; @Wu2004; @Saura2007; @Simova2012; @Arganaraz2014; @Frazier2021].
<!-- Advancing the science of scaling in landscape ecology is key to "permit the prediction and inference of quantities measured at one scale to another" [@Frazier2023] -->
<!-- also multi-scale is key to habitat assessment, see "Multi-scale habitat selection modeling: a review and outlook" -->
<!-- Predictable scaling relationships can be exploited to estimate spatial patterns at higher resolutions -->
Nonetheless, the effects of changing the sampling extent can be less predictable than those of changing grain size [@Wu2004].
For instance, in landscapes with small extent to grain ratio, boundary effects are likely to dominate and truncate patches resulting in unreliable metric computations.
In order to address scale issues, studies should explicitly consider the concept of scope, namely the ratio of extent to resolution [@Schneider2001].
This involves three fundamental steps, i) considering scope guidelines when designing experiments, ii) explicitly reporting grain and extent allowing quantitative scope comparisons with other studies, iii) and comparisons and replications based on scope [@Frazier2023].

<!-- TODO: section about landscape metric rescaling [@Saura2007; @Arganaraz2014; @Frazier2014; @Frazier2016; @Frazier2021]? -->

@@

<!-- TODO: mention hierarchichal patch dynamics [@Wu1999]? -->

<!-- (ii) choice of LULC scheme/thematic resolution and image processing/classification procedure: https://doi.org/10.1016/j.foreco.2009.12.016 https://doi.org/10.1016/j.isprsjprs.2007.03.004 https://doi.org/10.1007/s10021-005-0119-1 https://doi.org/10.1007/s10980-006-9035-9 https://doi.org/10.1007/s10980-022-01546-1 -->
<!-- Scale and heterogeneity are inherently linked (Wu 2007), so it is not surprising that the results indicate downscaling accuracy varies with heterogeneity. -->
<!-- TODO: sensitivity to thematic resolution [@Huang2006; @Bailey2007; @Buyantuyev2007; @Castilla2009]. -->

@@

<!-- MH: I think most of this is either not relevant here because not computational focused or already covered above -->

@@

Many landscape ecology studies are based on the raster data model.
However, the decision on the data model is often driven by data and software availability, or the familiarity with the approach [@Costanza2019].
For example, many LULC maps are provided as raster as they often related to underlying remote sensing products.
Yet, nowadays, vector data is increasingly available, and most geospatial software supports it.
